---
title: "Correction and Interpretation of stable Carbon isotopes"
output: html_notebook
---
  
#### Isotope Dendrochronology
#### COURSE: DISC, LTRR, UNIVERSITY OF ARIZONA. INSTRUCTORS: S. BELMECHERI & P. SZEJNER
  
### Instructions:
  
The following code and comments should further your experience to analyze and treat tree ring d13C data (including atmosphere and climate data). This exercise relies upon raw tree-ring d13C measurements, atmospheric d13C measurements, and instrumental climate data.

There are a few questions throughout to catalyze understanding and critical thinking. The analysis is broken into four main parts (including 17 questions). At the end, you should have gone through many of the steps commonly used for treating, correcting and, interpreting a d13C time series.

You will need to download the R project from https://github.com/SBelmecheri/DISCO-Notebook1.git and set your own working directory. Once you open the ```DISCO-Notebook1.Rproj``` in R studio you can open the ```RNotebook1.Rmd``` file to proceed with this exercise.

## PART 0: Setting up directories and files.

For the following exercise, you will need the following files in one directory:    
1. A file containing tree-ring d13C measurements. The filename is "LIL.txt".  
2. A file containing atmospheric 13C data. The filename is "ATM13C.txt".  
3. A file containing regional temperature data. The filename is "CRU_tt.txt".  
4. A file containing regional precipitation data. The filename is "CRU_PP.txt".  
4. A file containing regional Vapor pressure deficit data. The filename is "CRU_vpd.txt".  

```{r}
# check the directory with the following line.
# getwd()

# You should see the above file names (and anything else in this directory) when you type (or copy and paste) the following line:
list.files()
```

If you see these files listed, you're ready to roll!

## PART 1: Importing, plotting, and assessing raw tree-ring isotope data [d13C] & Atmospheric Data

ABOUT THE CHRONOLOGY [LIL, Pinar de Lillo]:   
* TIME: 1901-2002.   
* ELEVATION: 1600m a.s.l.   
* LATITUDE: 43.05.   
* LONGITUDE: -5.25.   
* SPECIES: Pinus sylvestris.  
* STAND STRUCTURE: Open forest.   
* ASPECT: NW.   
* SLOPE: 28 degrees.  

Import the tree ring d13C dataset:
```{r}
# You will also need to ensure that the name of the file is correct.
TR13 <- read.table("LIL.txt", header=T)

#convert to a time-series object. 
TR13.ts <- ts(TR13,start=1901, frequency=1)

#plot the raw data
ts.plot(TR13.ts,ylab="d13C (permill.VPDB)",xlab="Years", col="#9C964A")
```
Question 1: Observe the trend (directionality). What do you see?

Look more closely at various time windows:
```{r}
layout(matrix(1:2,nrow = 1, ncol = 2,byrow = T), widths = c(3,3),heights = c(3,3),respect = TRUE)
ts.plot(TR13.ts,ylab="d13C (permill.VPDB)",xlab="Years",xlim=c(1901,1960), col="#9C964A",lty=1, lwd=2)
ts.plot(TR13.ts,ylab="d13C (permill.VPDB)",xlab="Years",xlim=c(1960,2002), col="#9C964A",lty=3,lwd=2)
```

Question 2: Do you observe a difference between various time windows?

Question 3: Describe the difference between various time windows?

Import the atmospheric d13C dataset:
```{r}
ATM13 <- read.table("ATM13C.txt", header=T)

#convert to a time-series object
ATM13.ts <- ts(ATM13,start=1901,frequency=1)

#plot the atmospheric d13C dataset
ts.plot(ATM13.ts,ylab="d13C (permill.VPDB)",xlab="Years")
```

Question 4: Observe the trend (directionality) and interannual variability. What do you see?

```{r}
# plot the raw tree ring d13C dataset and the atmospheric d13C dataset
layout(matrix(1:2,nrow = 1, ncol = 2,byrow = T), widths = c(3,3),heights = c(3,3),respect = TRUE)
ts.plot(TR13.ts,ylab="d13C (permill.VPDB)",xlab="Years", col="#9C964A",main="TREE RINGS")
ts.plot(ATM13.ts,ylab="d13C (permill.VPDB)",xlab="Years", main="ATMOSPHERE")
```

Question 5: Describe the difference between absolute tree ring and atmospheric d13C values. What drives this difference?

Question 6: Compare both trends (directionality, magnitude). Describe the similarities/differences.

## PART 2: Correcting Tree Ring 13C raw Data
### Suess Effect
As a consequence of industrialization, the isotopic composition of atmospheric CO2 has changed.  
This change corresponds to a decline in atmospheric d13C related to burning of fossil fuels (enriched in 12C and/or low in 13C). This trend or effect is termed Suess Effect (Keeling, 1979), is recorded in tree-ring raw d13C data, and should therefore be removed before analyzing climate or ecophysiology of a tree ring d13C chronology.

This correction can be made as follows:
```{r}
TR13.corr <- TR13.ts -(ATM13.ts + 6.5)

# the 6.5  corresponds to the  pre industrial atmospheric d13C value.
# the correction factor (ATM13.ts + 6.5) is applied to express tree-ring d13C values relative
# to a pre-industrial standard value of 6.5 permill i.e. remove the trend of 13C that result 
# from a decline due to the Suess effect.

```

Add the corrected tree ring data to the plot with the raw data:
```{r}
ts.plot(TR13.ts,ylab="d13C (permill.VPDB)",xlab="Years", col="#9C964A")
lines (TR13.corr, col="#F21A00")
legend("bottomleft",c("d13C raw", "d13C Suess"),lwd= c(1,1), bty = "n", col=c("#9C964A","#F21A00"),
       text.col=c("#9C964A","#F21A00"),ncol=1)

# Note that you may need to adjust the limits of the Y axis to accommodate the values-range of raw and corrected 13C data. You can do that using ylim=c(,) to set min and max Y axis values.
```

Question 7: How does the corrected d13C compare to the raw d13C. 

### CO~2~ Effect
The correction of raw 13C tree ring from the Suess effect yields a d13C chronology that contains two signals: atmospheric CO2 and climate. These signals are related to tree physiological response to these environmental variables as was reviewed in the stable isotope theory lectures.

Tree-ring d13C chronologies can be corrected for changes in the carbon dioxide content of the atmosphere. This allows analysis and interpretation of d13C variations and trends in response to climate variability/trends only.

This correction, termed PIN correction (for pre-industrial) corresponds to a statistical procedure that calculates the d13C values that would have been obtained under pre-industrial conditions (pre-increase in atmospheric CO2). This procedure has been published by McCarroll et al. 2008 [https://doi.org/10.1016/j.gca.2008.11.041]

Apply the PIN correction to the raw d13C data:
```{r}
# First install the PIN R package (change the pathway to your folder)
install.packages(pkgs="pin_0.5", repos = NULL, type = "source")
library(pin)

```

the pin function is as follows:

pin(year, d13c, co2, alpha = 0.9, firstyear =, lastyear =)

year = 1D array of years. The years corresponding to the d13C (tree ring) measurements.

d13c = vector of atmospherically corrected d13 C values.

co2  = smoothed atmospheric co2 record (e.g. from mauna loa).

alpha =	the span value used by the loess routine.

firstyear= 	firstyear / lastyear allows selecting a subset of the data.


for co2 data, the Mauna Loa data from Keeling are included as a proxy in this package.

However, these data have been updated and the most recent data are included in this Rproject/Notebook (Belmecheri & Lavergne 2021):
```{r}
#read/import atmospheric CO2 data and convert them to a time serie
co2 <- ts(read.table("ATMCO2.txt", header=T),start=1901, frequency=1)
```

Apply the pin correction:

After running this line,  yo will have to go to the console in R  and  press  
```Hit <Return> to see next plot```
```{r}
  
TR13.corr.pin <- pin(c(1901:2002), TR13.corr, co2, alpha = 0.9, firstyear =1901, lastyear =2002)
# Note: The function generates plots for the various steps of the PIN calculation. You will need to hit return multiple times. Looking carefully at the plots is advised as it demonstrates the theoretical assumptions and bounds of tree physiological responses to increases in atmospheric CO2 concentrations.
```


Add the pin corrected tree ring data to the plot with the raw data and Suess corrected data:
```{r}
ts.plot(TR13.ts,ylab="d13C (permill.VPDB)",xlab="Years", col="#9C964A")
lines(TR13.corr, col="#F21A00")
lines(TR13.corr.pin, col="#E1AF00", lwd=1.3)
legend("bottomleft",c("d13C raw", "d13C Suess", "d13C PIN"),lwd= c(1,1,1.3),bty = "n", col=c("#9C964A","#F21A00","#E1AF00"),
       text.col=c("#9C964A","#F21A00","#E1AF00"),ncol=1)

# Note that you may need to adjust the limits of the Y axis to accommodate the values-range of raw and corrected 13C data. You can do that using ylim=c(,) to set min and max Y axis values.

```

Question 8: How does the PIN-corrected d13C compare to the Suess-corrected and raw d13C?

Question 9: Do you believe the differences between raw and corrected (SUESS, PIN) d13C are significant?

## PART 3: Importing instrumental data and computing correlations with tree ring chronologies

### Climatology of the growing season 
climate data are from the Climatic Research Unit (CRU) for the grid point near the tree-ring data:
```{r}
# import gridded temperature data & Convert to a time series 
gridded.tt.data <- ts(read.table("CRU_TT.txt",header=T),start=1901,frequency=1)

# import gridded precipitation data & Convert to a time series 
gridded.pp.data <- ts(read.table("CRU_PP.txt",header=T),start=1901,frequency=1)

# import gridded VPD data & Convert to a time series 
gridded.vpd.data <- ts(read.table("CRU_VPD.txt",header=F, skip=3),start=1901,frequency=1)

```

Explore the climate data (seasonality, trends) for all climate parameters. Below is an example for Temperature:
```{r}
# plot temperatures for all months with a legend
ts.plot(gridded.tt.data,col=1:8,lty=1:3,xlim=c(1870,2002))
legend("topleft",legend=colnames(gridded.tt.data),col=1:8,lty=1:3,bty = "n")

```

```{r}
# Calculate and plot temperature seasonality/climatology
avg.month.tt <- colMeans(window(gridded.tt.data,start=1901,end=2002))
#Plot the mean monthly temperatures for the 
barplot(avg.month.tt,las=3,ylim=c(0,20),ylab="Deg C",main="Average Monthly temperatures for LIL")

```

Question 10: Do you observe a trend in the temperature, precipitation and VPD records?

Question 11: Looking at the seasonality, describe the growing season conditions?

```{r}
# repeat these steps for precipitation and VPD.
```

### Correlation between tree ring d13C and climate data

combine (here a multiple time series object) the tree ring d13C and instrumental data in the same data matrix for correlations.

First, start with d13C corrected for the Suess effect:
```{r}
TR13.corr_tt <- ts.union(TR13.corr,gridded.tt.data) # temperature; 
colnames(TR13.corr_tt) <- c("TR13.corr", colnames(gridded.tt.data))
TR13.corr_pp <- ts.union(TR13.corr,gridded.pp.data) # precipitation; 
colnames(TR13.corr_pp) <- c("TR13.corr", colnames(gridded.tt.data))
TR13.corr_vpd <- ts.union(TR13.corr,gridded.vpd.data) # vapor pressure deficit
colnames(TR13.corr_vpd) <- c("TR13.corr", colnames(gridded.vpd.data))
```
Compute a correlation matrix for temperature:
```{r}
TR13.corr_tt.correl <- cor(TR13.corr_tt,use="p")
```
Look at the correlation matrix:
```{r}
TR13.corr_tt.correl
```
or look at just the first 2 significant digits:
```{r}
round(TR13.corr_tt.correl,2)
```

compute correlation matrices for precipitation and VPD:
```{r}
TR13.corr_pp.correl <- cor(TR13.corr_pp,use="p")
TR13.corr_vpd.correl <- cor(TR13.corr_vpd,use="p")
```

and plot the correlations between monthly climate data and the d13C corrected for the Suess effect chronology. This will be plotted for the 3 climate variables:
```{r}
layout(matrix(1:3,nrow = 1, ncol = 3,byrow = T), widths = c(4,4,4),heights = c(3,3,3),respect = TRUE)
#Temperature
barplot(TR13.corr_tt.correl[1,2:13],las=3,col=c("yellow"),ylab="Correlation", main="TEMP", ylim=c(-0.5,0.5))
#precipitation
barplot(TR13.corr_pp.correl[1,2:13],las=3,col=c("blue2"),ylab="Correlation", main="PCP", ylim=c(-0.5,0.5))
# VPD
barplot(TR13.corr_vpd.correl[1,2:13],las=3,col=c("yellow"),ylab="Correlation", main="VPD", ylim=c(-0.5,0.5))
```

Question 12: Describe the correlations between d13C and various months of the year/growing season.

Question 13: Do you observe similarities/differences in the d13C response to the seasonality of the climate variables?

Tree growth and physiology respond to climate variation often over several months.Therefore, when investigating the climate signal recorded in d13C or reconstructing climate, we often try to better understand climate response by checking correlations over a wide variety of seasons. This is most easily done by averaging months together and computing correlations.  

An example with March through June temperatures is shown below using the ```"rowMeans()"``` function. This function removes the time-series information so we need to add this back with the ```"ts()"``` function.

Below is a two-steps example for overwriting the initial variable (tt.Mar_Jun)  instead of defining a new one.
```{r}
tt.Mar_Jun <- rowMeans(gridded.tt.data[,c(3,4,5,6)])
tt.Mar_Jun <- ts(tt.Mar_Jun,start=1901,frequency =1)
```


New seasonal windows can then be created for Winter, Spring, summer (June-July-August), or a growing season (e.g. April-September):
```{r}
# Here create multiple seasonal windows for the climate parameters. 
```

We can then combine this new seasons to our d13C and monthly temperature matrix. The example below can be enhanced with more season developed in the previous step:
```{r}
TR13.corr_tt.seasons <- ts.union(TR13.corr_tt, tt.Mar_Jun)
```

compute a correlation matrix for temperature with monthly and seasonal data; and rename the columns to something nicer (add more column names corresponding to the seasons computed above):
```{r}
TR13.corr_tt.seasons.correl <- cor(TR13.corr_tt.seasons,use="p")
colnames(TR13.corr_tt.seasons.correl) <- c(colnames(TR13.corr_tt),"Mar_Jun")
```

and plot the correlations for the individual months and the seasons you have chosen:
```{r}
barplot(TR13.corr_tt.seasons.correl[1,2:15],las=3,col=c("goldenrod2"),ylab="Correlation")
```
Compute seasonal correlations for precipitation and VPD:
```{r}
# repeat the steps above (combine monthly and seasonal climate-precipitation and vpd with TR13)
```

Compute correlations with d13C PIN corrected. Use climate matrices with monthly and seasonal windows and combine them to d13C PIN corrected:
```{r}
# below is an example for temperature
TR13.corr.pin_tt.seasons <- ts.union(TR13.corr.pin, TR13.corr_tt,tt.Mar_Jun) 
# this needs to include more seasonal windows from previous steps
TR13.corr.pin_tt.seasons.correl <- cor(TR13.corr.pin_tt.seasons,use="p")
# do the same for precipitation and vpd
```

### Suess vs. PIN corrections
Compare the correlations coefficient and seasonality between Suess and PIn d13C:
```{r}
# you can use a correlation matrix plot
# Or a barplot (this can be a grouped barplot using the argument beside=T).

```

Question 14: What do you observe? are there differences, improvement of the correlation coefficients? a change in the significant months?

Question 15: Which correction would you use for a climate reconstruction.

Question 16: Which correction would you use for an ecophysiological study.

## PART 4: Isotopic Discrimination as means to correct d13C

We have seen above that the d13C time series can be treated to remove the effect of declining atmospheric d13C and the effect of increasing atmospheric CO2 concentration (PIN). The Suess Correction is straightforward and necessary as it avoids artifactual trends in tree ring d13C. 

The PIN correction is a statistical treatment to remove the response of tree physiology to CO2, consequently it assumes that this response is bound by two defined stomatal conductance adjustments.

Another way to treat the d13C data is to calculate isotopic discrimination. This procedure takes care of the Suess effect and allows interpreting variation of the isotopic fractionation in term of photosynthesis processes and  availability of leaf internal CO2.

Calculate the isotopic discrimination:
```{r}

d13C.disc <- (ATM13.ts-TR13.ts)/(1+TR13.ts/1000)

# the above is the full & Appropriate equation. It is not unusual to see a simplified version
# consisting of the numerator term (ATM13.ts-TR13.ts) used in tree ring publications.
# However, it is strongly suggested to use the full equation as in Farguhar et al., 1982
```

Plot isotopic discrimination along with d13C corrected from the suess effect:
```{r}
plot.ts(TR13.corr, ylab="d13C (permil.VPDB)",xlab="Years", col="#9C964A")
par(new=T)
plot.ts(d13C.disc,axes=F, xlab="", ylab="", col="gold4",lty=3,lwd=4)
axis(4, at=seq(15,19 , by=1))
mtext(side= 4, text = "D13C (permil.VPDB)", line = 2.5)

```

Question 17: What is the difference between the d13C (tree rings d13C corrected from Suess effect) and D13C (isotopic discrimination). Describe the observed trends in terms of isotopic fractionation and associated physiological responses.

Use the climate correlations investigated above to discuss how tree physiology responds to changes in trends and variability of the climate factors and atmospheric CO2 concentration.

